The very dependent recursive structure of iterated parametricity in indexed form

### 2025-12-29 paper/paper.tex
Bib item names

Introduced by Reynolds in the context of programming language semantics~\cite{Reynolds}, relational parametricity happens to be connected to several similar concepts in logic and mathematics. Associating binary propositional relations to types in its original formulation, it can also associate proof-relevant relations~\cite{GhaniNordvallOrsanigo16} or unary relations, that is predicates (proof-irrelevant case) or families (relevant case). These unary or binary, relevant or proof-irrelevant relations express what it means to belong in a type (unary case) or to be equal in a type (binary case) from an observational point of view. For instance, the canonical predicate associated to $A \rightarrow B$ is the predicate over functions $f$ in $A \rightarrow B$ such that, for all $a$ in the canonical predicate associated to $A$, $f(a)$ is in the canonical predicate associated to $B$. And, similarly, the canonical relation associated to $A \rightarrow B$ is the relations over functions $f$ and $f'$ in $A \rightarrow B$ such that, for all $a$ and $a'$ related through the canonical relation associated to $A$, $f(a)$and $f'(a')$ are related through the canonical relation associated to $B$.

@@

The internalisation of parametricity in unary indexed form was studied by \cite{Moulin,Bernardy} directly leading to a type theory of iterated parametricity. Such type theory could be interpreted in a ``unary'' form of cubical set~\cite{BernardyCoquandMoulin}, thus implicitly in fibered form since, in a binary cubical set (resp. in a unary cubical set and simplicial set as well), a set of segments over two points, that is a relevant relation, is given by first and second projection rather than as a function to $\MySet$ (resp. a family is given as a \emph{total space} projected over a set of points $A:\MySet$ rather than a family of type $A \rightarrow \MySet$. This connection between iterated binary parametricity in fibered form and cubical sets includes an identification of degeneracies and reflexivity proofs, which can also be found in Johann and Sojakova~\cite{JohannSojakova}. In particular, it became clear at this point that iterated binary parametricity (without reflexivity) produces semi-cubical sets while it was also observed\footnote{The authors learned it from Hugo Moeneclaye and Thorsten Altenkirch.} that iterated unary parametricity produces augmented semi-simplicial sets.

>>>>

Introduced by Reynolds in the context of programming language semantics~\cite{reynolds83}, relational parametricity happens to be connected to several similar concepts in logic and mathematics. Associating binary propositional relations to types in its original formulation, it can also associate proof-relevant relations~\cite{GhaniNordvallOrsanigo16} or unary relations, that is predicates (proof-irrelevant case) or families (relevant case). These unary or binary, relevant or proof-irrelevant relations express what it means to belong in a type (unary case) or to be equal in a type (binary case) from an observational point of view. For instance, the canonical predicate associated to $A \rightarrow B$ is the predicate over functions $f$ in $A \rightarrow B$ such that, for all $a$ in the canonical predicate associated to $A$, $f(a)$ is in the canonical predicate associated to $B$. And, similarly, the canonical relation associated to $A \rightarrow B$ is the relations over functions $f$ and $f'$ in $A \rightarrow B$ such that, for all $a$ and $a'$ related through the canonical relation associated to $A$, $f(a)$and $f'(a')$ are related through the canonical relation associated to $B$.

@@

The internalisation of parametricity in unary indexed form was studied by \cite{Moulin,Bernardy} directly leading to a type theory of iterated parametricity. Such type theory could be interpreted in a ``unary'' form of cubical set~\cite{bernardy15}, thus implicitly in fibered form since, in a binary cubical set (resp. in a unary cubical set and simplicial set as well), a set of segments over two points, that is a relevant relation, is given by first and second projection rather than as a function to $\MySet$ (resp. a family is given as a \emph{total space} projected over a set of points $A:\MySet$ rather than a family of type $A \rightarrow \MySet$. This connection between iterated binary parametricity in fibered form and cubical sets includes an identification of degeneracies and reflexivity proofs, which can also be found in Johann and Sojakova~\cite{johann17}. In particular, it became clear at this point that iterated binary parametricity (without reflexivity) produces semi-cubical sets while it was also observed\footnote{The authors learned it from Hugo Moeneclaye and Thorsten Altenkirch.} that iterated unary parametricity produces augmented semi-simplicial sets.
